---
title: "StatTeam6_final"
author: "Stuart Gavidia, Joana Hadzhilazova, Eric Aguirre, & Lisa Ha"
date: "12/7/2021"
output: html_document
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
knitr::opts_chunk$set(message = FALSE)
# Load Packages
library(gapminder)
library(openintro)
library(ggthemes)
library(dplyr)
library(stat242)
library(moderndive)
library(infer)
library(ggplot2)
library(tidyr)
library(kableExtra)
library(patchwork) # For aesthetics

# Load data 
raw_data <- stat242::project_data
```


Part I - Lisa

# Exploratory Data Analysis

We begin our study of the `project_data` data with some exploratory data analysis. This particular data set has 8840 rows and 8 columns. The variables are `country`, `continent`, `year`, `life_exp`, `population`, `num_phones`, `net_users` and `non_net_users`. The first few are fairly self-explanatory as to what they represent. For the others, `life_exp` is the life expectancy of someone born in that country during that specific year, and `population` is the population of the country in the given year.

For This presentation, we will be focusing on the relationship between life expectancy and the number of internet users in a given country. First, let's examine net_users by continent:


```{r}
ggplot(raw_data, aes(net_users, continent))+
  geom_boxplot(color = IMSCOL["blue", "full"])+
  theme_minimal()+
  stat_summary(fun = mean, geom = "point", col = IMSCOL["red", "full"])+
  stat_summary(fun = mean, geom = "text", col = IMSCOL["red", "full"],
               position = position_nudge(x = 5000, y = 0.5), 
               aes(label = paste("Mean: ", round(..x.., digits = 2))))+
  labs(
    title = "Net Users", 
    subtitle = "by continent, 1979 - 2018", 
    x = "Net Users",
    y = "Continent", 
    caption = "Source: Project Data"
  ) 
```

Let's remove some of the outliers to make this a bit easier to read:

```{r}
ggplot(raw_data, aes(net_users, continent))+
  geom_boxplot(color = IMSCOL["blue", "full"])+
  theme_minimal()+
  stat_summary(fun = mean, geom = "point", col = IMSCOL["red", "full"])+
  stat_summary(fun = mean, geom = "text", col = IMSCOL["red", "full"],
               position = position_nudge(x = 5000, y = 0.5), 
               aes(label = paste("Mean: ", round(..x.., digits = 2))))+
  labs(
    title = "Net Users", 
    subtitle = "by continent, 1979 - 2018", 
    x = "Net Users",
    y = "Continent", 
    caption = "Source: Project Data"
  ) +
  xlim(1,200000000)
```

It's quite clear even after removing a good number of the outlying data, that there are many extreme values across all continents, and most of the data lie on the lower side of the spectrum.

Now, let's take a look at life_exp:

```{r}
ggplot(raw_data, aes(life_exp, continent))+
  geom_boxplot(color = IMSCOL["blue", "full"])+
  theme_minimal()+
  stat_summary(fun = mean, geom = "point", col = IMSCOL["red", "full"])+
  stat_summary(fun = mean, geom = "text", col = IMSCOL["red", "full"], 
               position = position_nudge(x = 0, y = 0.5), 
               aes(label = paste("Mean: ", round(..x.., digits = 2))))+
  labs(
    title = "Life Expectancy", 
    subtitle = "by continent, 1979 - 2018", 
    x = "Life Expectancy",
    y = "Continent", 
    caption = "Source: Project Data"
  )
```

Overall there is a lot less variation compared to the data for net_users. Most continents have a mean life expectancy hovering around the high 60s/low 70s, apart from africa with a mean of about 58.

Part II -Eric

# Data Visualization 

Now, let's go through a few plots to provide some context.

### A Scatterplot of net_users v year

```{r}
ggplot(project_data, aes(year, net_users, color = as.factor(continent)))+
  theme_clean()+
  geom_point()
```
The scatter plot shows the sharpest increase of internet users in Asia, starting around 2005. There is also a dramatic increase in North America, which started much earlier, around 1995. Other continents, however, only showed a steady increase of users beginning in the late 90s.

### Extra Scatterplot life_exp v year

```{r}
ggplot(project_data, aes(year, life_exp, color = as.factor(continent))) +
  theme_clean()+
  geom_point()
```
This plot shows a very steady increase in life expectancy across all continents between 1979 and 2018. From the distribution you can tell that Africa has almost exclusively the lowest life expectancy, while the other continents tend to vary a bit more.

### Life Expectancy v net users
Now, let's combine the two graphs, again coloring by continent:

```{r}
ggplot(project_data, aes(net_users, life_exp, color = as.factor(continent))) +
  theme_clean()+
  geom_point()
```
While most of the data are consolidated to the extreme left of the scale, some correlation can be seen, especially in North America and Asia.

### Separated by continent

Let's look at this data again, separated by continent.

```{r}

# filter by continent
asia <- project_data %>%
  filter(continent=="Asia")
africa <- project_data %>%
  filter(continent=="Africa")
europe <- project_data %>%
  filter(continent=="Europe")
s_amer  <- project_data %>%
  filter(continent=="South America")
n_amer <- project_data %>%
  filter(continent=="North America")
oceania <- project_data %>%
  filter(continent=="Oceania")


# create plots for each continent
asia_plot <- ggplot(asia, aes(net_users, life_exp))+
  geom_point(color = "#C59900")+
  labs(
    title = "Asia",
    x = "Internet Users (millions)",
    y = "Life Expectancy"
  )+
  theme(axis.text.x = element_text(angle = 45))


africa_plot <- ggplot(africa, aes(net_users, life_exp))+
  geom_point(color = "#F8766D")+
  labs(
    title = "Africa",
    x = "Internet Users (millions)",
    y = "Life Expectancy"
  )+
  theme(axis.text.x = element_text(angle = 45))

euro_plot <- ggplot(europe, aes(net_users, life_exp))+
  geom_point(color = "#39B600")+
  labs(
    title = "Europe",
    x = "Internet Users (millions)",
    y = "Life Expectancy"
  )+
  theme(axis.text.x = element_text(angle = 45))

s_am_plot <- ggplot(s_amer, aes(net_users, life_exp))+
  geom_point(color = "#FF61C9")+
  labs(
    title = "S America",
    x = "Internet Users (millions)",
    y = "Life Expectancy"
  )+
  theme(axis.text.x = element_text(angle = 45))

n_am_plot <- ggplot(n_amer, aes(net_users, life_exp))+
  geom_point(color = "#00C0B8")+
  labs(
    title = "N America",
    x = "Internet Users (millions)",
    y = "Life Expectancy"
  )+
  theme(axis.text.x = element_text(angle = 45))

ocea_plot <- ggplot(oceania, aes(net_users, life_exp))+
  geom_point(color = "#00ABFD")+
  labs(
    title = "Oceania",
    x = "Internet Users (millions)",
    y = "Life Expectancy"
  )+
  theme(axis.text.x = element_text(angle = 45))

# combine plots
asia_plot + africa_plot + euro_plot + s_am_plot + n_am_plot + ocea_plot+
  plot_annotation(title = "Life Expectancy v. Internet Users by Continent",
                  subtitle = "1979 - 2018",
                  caption = "Source: Gapminder")
```
The distinction by continent helps us more clearly see the correlation of the two variables, despite data consolidation on the left.


Part III - Joana

# Linear Modeling

## The focus here will be comparing Life Expectancy and Net Users Globally and the relationship between these two variables will be analyzed using a scatterplot.

### Build model 

```{r}
life_net_mod <- lm(formula = life_exp ~ net_users, 
                   data = project_data)
```

### Get the regression table

```{r}
reg_table <- tibble(get_regression_table(life_net_mod))
```

### Plot regression model

```{r}

ggplot(project_data, aes(net_users, life_exp, size = population/10^6))+
  geom_point(alpha = 0.2)+
  geom_smooth(method = "lm", se = FALSE, show.legend = FALSE)+
  labs(
    title = "Life Expectancy v. Net Users Globally",
    subtitle = "1979 - 2018",
    size = "Population (millions)",
    x = "Net Users",
    y = "Life Expectancy",
    caption = "Source: Stat242"
  )+
  theme_economist()+ # from ggthemes
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1))

```

The scatterplot indicates that there is a positive linear relationship between Life Expectancy and the number of Net Users Globally. Visually, it appears that as the life expectancy of the population (in millions) increases, the number of net users increases as well. Additionally, since the number of net users are recorded globally, it is important to specify which regions or continents we are analyzing. The continents in question include Africa, Asia, Europe, North America, Oceania, and South America. A correlation (Cor) value has been recorded below for each continent respectively to portray the (strength of the) relationship between the two variables in question.


```{r}
get_correlation(data = project_data %>% group_by(continent),
                formula = life_exp[!is.na(life_exp)] ~ net_users[!is.na(net_users)])


```

As we can see from the table, Oceania has the highest correlation and Asia has the lowest correlation. Additionally, Europe and North America have almost identical correlations. Similarly, Africa and South America have very similar correlations.
 
Furthermore, it is worth considering how the presence of outliers influences the correlation of a continent. From the graph above we can see that there is an outlier, as indicated by the small cluster of dots below and toward the middle of the regression line. We know that visually, the size of the points is the population and can assume that the outliers are the same country. The presence of the outlier(s) likely contributes to the significantly lower correlation of Asia compared to the other continents, and finding which country in Asia corresponds to the outlier, and consequently removing it, can increase the correlation of Asia. Analysis of a facet wrap graph can help with this task.

The time frame throughout which our data was recorded is the span of years from 1979 to 2018 as seen in the facet wrap graph below. 

```{r}

ggplot(project_data, aes(net_users, life_exp, size = population/10^6))+
  geom_point(alpha = 0.2)+
  geom_smooth(method = "lm", se = FALSE, show.legend = FALSE)+
  facet_wrap(~year)+
  labs(
    title = "Life Expectancy v. Net Users Globally",
    subtitle = "1979 - 2018",
    size = "Population (millions)",
    x = "Net Users",
    y = "Life Expectancy",
    caption = "Source: Stat242"
  )+
  theme_economist()+ # from ggthemes
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1))

```


Zooming in on the facet wrap graph we can see that the three-dot-cluster that make up the outlier, appear in the years 2016-2018. Furthermore, the size of the dot under these three years corresponds to a small population according to the key at the top of the graph. Looking at the number of net user associated with the outlier is approximately 4.5 x 108.

Knowing this information, we can find which country in the project_data table contains these parameters and can identify which country is the outlier.


```{r}

View(project_data)

```


We conclude that the outlier is India. It is tabulated under in the year 2018, is part of the continent Asia, and contains 4.71 x 108 net users. This confirms the descriptions we observed earlier from the facet wrap graph.
 
We can now write code to exclude India from the dataset and recalculate the correlation of Asia.


```{r}
project_data_2 <- project_data %>% 
  filter(country != "India")

get_correlation(project_data_2, life_exp ~ net_users,na.rm=TRUE)

```


The correlation of Asia is now higher by approximately 0.098 units than the original as observed in the table above. There is now a stronger relationship between Life Expectancy and number of Net Users in Asia.
 
 
Finally, recreating the scatterplot of our data with the new correlation lacking India, we see that the points are more closely clustered around the regression line and the outlier dot(s) are missing. The confirmation that there is a relationship between Life Expectancy and Net Users Globally stands.


```{r}

ggplot(project_data_2, aes(net_users, life_exp, size = population/10^6))+
  geom_point(alpha = 0.2)+
  geom_smooth(method = "lm", se = FALSE, show.legend = FALSE)+
  labs(
    title = "Life Expectancy v. Net Users Globally",
    subtitle = "1979 - 2018",
    size = "Population (millions)",
    x = "Net Users",
    y = "Life Expectancy",
    caption = "Source: Stat242"
  )+
  theme_economist()+ # from ggthemes
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1))

```



Part IV - Stuart

### Prediction using model from part 3
## What would life expectancy be in a South American country in the year 6000?

We wanted to look at what the expected life expectancy would be in the year 6000, based on the linear model we could create with the current data.

```{r}

pred <- linear_model_prediction(life_net_mod, 6000)

```

## Test: is this significantly  lower from the average?

```{r}
test_stat <- numeric_z(
  x = pred,
  mu = mean(s_amer$life_exp),
  s = sd(s_amer$life_exp)
)

pnorm(test_stat, lower.tail = TRUE)
```

Our p-value is greater than 0.05, we fail to reject the null hypothesis. Therefore we conclude, a South American country in the year 6000 will not have an unusually low life expectancy.

## Proportion net users of a country with rest of world 

We wanted to look at a country in North America, and compare the proportion of internet users with the rest of the globe. Is the proportion in Mexico with internet users over 1000 in the year 2015 significantly different than the proportion of internet users with the rest of the globe?

```{r}
group_test_data <- project_data %>% 
  filter(year == 2015 & !is.na(net_users)) %>%
  mutate(is_greater_1000 = 
           case_when(
             net_users >= 1000 ~ 1,
             net_users < 1000 ~ 0
           ),
         samp =
           case_when(
             country == "Mexico" ~ "Mexico",
             TRUE ~ "Global"
           )
  ) %>% 
  select(samp, is_greater_1000) %>% 
  group_by(samp) %>% 
  summarize(count = n(),
            users_greater_1000 = sum(is_greater_1000)/count)

mexico_users <- group_test_data %>% 
  filter(samp == "Mexico")

global_users <- group_test_data %>% 
  filter(samp != "Mexico")

test_stat_2 <- categorical_z(
  p_hat = mexico_users$users_greater_1000,
  p = global_users$users_greater_1000,
  n = mexico_users$count
)

pnorm(test_stat_2, lower.tail = FALSE)
```

The p-value is greater than 0.05 so we fail to reject the null hypothesis. Therefore we conclude, the proportion of net users in Mexico with net users above 1000 in the year 2015 is significantly different from the rest of the globe.
